---
title: "Hassle-free HTML tables with htmltable"
author: "Christian Rubba"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## htmltable
HTML tables are a standard way to display tabular information on the internet. Parsing and transforming HTML tables into R objects is very straightforward with the `readHTMLTable()` function of the *XML* package. But tables on the web are primarily designed for display and consumption of data, and not for analytical purposes. Peculiar design choices for HTML tables are thus oftentimes made which tend to produce useless outputs when run through `readHTMLTable()`. I found that sometimes these outputs could be saved with a little bit of (tedious) post-processing, but oftentimes they could not. To make working with HTML tables easier and less time-consuming, I developed *htmltable*, a package for the R system that tries to alleviate these problems directly in the parsing stage when structural information are still available. Its main advantages over `readHTMLTable()` are twofold:

- Consideration of row and column spans in HTML table body and header cells
- More control over the process that translates HTML cells into R table cells

This vignette discusses the application of *htmltable* for two use cases where the package provides a significant improvement over `readHTMLTable()`. 

(This vignette uses the R packages *magrittr*, *tidyr* and *stringr* to process table outputs. Neither of the three is required for running *htmltable*.)

## Installation
You can install the package from the [GitHub](https://github.com/crubba/) repository:

```{r, eval = FALSE, echo = TRUE, results = "hide", warning = F, message = F}
devtools::install_github("crubba/htmltable")
library(htmltable)
```

## How to read HTML tables with htmltable()
The principle function of *htmltable* is (surprise!) `htmltable()`. The behaviour of `htmltable()` is very closely modelled after `readHTMLtable()`'s, and most argument names are the same. Any function call requires passing a value to its _doc_ argument. This may be one of two kinds:

1. a URL or file path for the HTML page where the table lives
2. a parsed HTML object of the entire page

Both methods return to you a single R table object. Unlike `readHTMLTable()`, `htmltable()` requires users to be specific about the table they would like to have returned. Both also require to set the _which_ argument to specify the exact table. This may be either a numeric value for the table's position in the page, or a character value that describes an XPath statement.

```{r, echo = FALSE, results = "hide", warning = F, message = F}
library(htmltable)
library(magrittr)
library(tidyr)
library(XML)
```


## 1. Corrections for rowspans and colspans by default
In many HTML tables, spans are used to allow cell values to extend over multiple cells. `htmltable()` recognizes spans and expands these tables automatically. To illustrate this feature, take a look at the HTML table in the Language section on this [Wikipedia page about Demography in the UK](en.wikipedia.org/wiki/Demography_of_the_United_Kingdom.html#Languages). Header information span over three consecutive rows. To get this table into R, we have to pass an identifiying information to the _which_ argument. Here, I use an XPath statement that I came up with during exploration of the HTML page with Web Developer Tools. One that works here is "//th[text() = 'Ability']/ancestor::table":

```{r}
url <- "http://en.wikipedia.org/wiki/Demography_of_the_United_Kingdom"
ukLang <- htmltable(doc = url, which = "//th[text() = 'Ability']/ancestor::table")
head(ukLang)
```

The header information have been recast into a format that respects the hierarchical order of the variables and yet only spans a single line in the R table. If you prefer a different seperator between variables, pass it to the _headerSep_ argument. This format was mainly chosen to make further processing of the table really easy. For example, using functionality from the *tidyr* package, the next couple of data cleaning steps may be the following:

```{r, fig.show='hold', results = 'hide', warning = F, message = F}
library(tidyr)
library(magrittr)

ukLang %<>% gather(key, value, -Ability)
```

This gathers the variables into a more useful long format. From this we can separate the variables using an appropriate regular expression such as " >> ". 

```{r, fig.show='hold'}
ukLang %>% separate(key, into = c("region", "language", "statistic"), sep = " >> ") %>% head
```

`htmltable()` also automatically expands row and column spans when they appear in the table's body. 

## 2. More control over cell value conversion
With `htmltable()` you have fine control over what part of the HTML table is used in the R table. You can exert this control through `htmltables()`'s _body_, _header_, _bodyFun_, _headerFun_, _rm\_escape_, _rm\_footnote_ and _rm\_superscript_ arguments. 

### _body_ and _header_ arguments
It is not possible for `htmltable()` to correctly identify header and body elements in all the tables. Although there is a sementically *correct* way to organize header and body elements HTML tables, web designers do not necessarily need to adhere to them to produce visually appealing tables. The *htmltable* package employs reasonable heuristics for identification but they are no garuantee. If you find that the table is not correctly assembled you can try to give the function more information through its _header_ and _body_ arguments. These arguments are used to pass information about which rows should be used for the contruction of the header and the body. Both accept numeric values for the rows, but a more robust way is to use an XPath that identifies the respective rows. To illustrate, take a look at this [Wikipedia page about the New Zealand General Election in 2002](http://en.wikipedia.org/wiki/New_Zealand_general_election,_2002#Electorate_results). This table uses in-table all columns spanning cells to classify General and Maori electorates (yellow background). This is a problem that we need to control for explicitly in the assembling stage. Here, I pass the XPath "tr[./td[not(@colspan = '10')]]" to the _body_ argument to explicitly discard all rows from the body that have a \<td\> cell with a colspan attribute of 10:

```{r, fig.show='hold'}
url <- "http://en.wikipedia.org/wiki/New_Zealand_general_election,_2002"
xp <- "//caption[starts-with(text(), 'Electorate results')]/ancestor::table"
nz1 <- htmltable(doc = url, which = xp, body = "tr[./td[not(@colspan = '10')]]")
head(nz1)
```

You might object that ideally these rows should not be discarded, but used for what they are -- variable/header information! The capability to process in-table variables is an issue that I leave to be solved in a future version of this package.

### Removal of unneeded information
Many HTML tables include additional information which are of little interest to data analysts. These are information encoded in superscript and footnote tags, as well as escape sequences. By default, `htmltable()` removes information from the first two and replaces all escape sequences by a whitespace. You can change this behaviour through the _rm\_superscript_, _rm\_footnotes_ and _rm\_escape_ arguments. 

## Conclusion
HTML tables are a valuable data source but they oftentimes violate even basic principles of data well-formedness. This is usually for good reasons since their primary purpose is to increases readability of tabular information. *htmltable*'s goal is to reduce the need for users to interfere when working with HTML tables by using inherent structural information as well as making some assumptions about the table's design. However, you are free to exert more control about the transformation by specifying the function's arguments. 


